---
title: "Week 2 Assignment - Leaflet map"
date: "1/8/2017"
output: html_document
---

```{r setup, include=FALSE}
library(leaflet)
library(dplyr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```
First, run code to download data of country centerpoints and populations, and then merge these into one data frame. As the former was using 2-letter country codes and the latter 3-letter ones, a third file was merged to map from one abbreviation to another. Finally, countries were divided into 4 quantiles by population, and a color was stored in the data frame to represent the quantile.
```{r echo = FALSE}
# coordinates
destfile <- "country_centroids_primary.csv"
if (!file.exists(destfile)) {
  download.file("http://gothos.info/resource_files/country_centroids.zip",
                "country_centroids.zip",
                method="curl")
  unzip("country_centroids.zip", c(destfile))
}
country_centers <- read.delim(destfile,
                              header = TRUE,
                              sep = "\t",
                              quote = "\"",
                              dec = ".",
                              fill = TRUE,
                              comment.char = "")

# country code mapping
destfile <- "country_codes.csv"
if (!file.exists(destfile)) {
  download.file("https://commondatastorage.googleapis.com/ckannet-storage/2011-11-25T132653/iso_3166_2_countries.csv",
                "country_codes.csv",
                method="curl")
}
country_codes <- read.table("country_codes.csv",
                            sep=",",
                            header = TRUE,
                            quote = "\"",
                            fill = TRUE)
countries <- merge(country_centers[, c("LAT", "LONG", "SHORT_NAME", "ISO3136")],
                   country_codes[, c("ISO.3166.1.3.Letter.Code", "ISO.3166.1.2.Letter.Code")],
                   by.x = "ISO3136",
                   by.y = "ISO.3166.1.2.Letter.Code")

# population data
destfile <- "API_SP.POP.TOTL_DS2_en_csv_v2.csv"
if (!file.exists(destfile)) {
  download.file("http://api.worldbank.org/v2/en/indicator/SP.POP.TOTL?downloadformat=csv",
                "populations.zip",
                method="curl")
  unzip("populations.zip", c(destfile))
}
country_populations <- read.table("API_SP.POP.TOTL_DS2_en_csv_v2.csv",
                                  sep=",",
                                  header = TRUE,
                                  fill = TRUE,
                                  skip = 4)
countries <- merge(countries,
                   country_populations[, c("Country.Code", "X2015")],
                   by.x = "ISO.3166.1.3.Letter.Code",
                   by.y = "Country.Code")
colnames(countries) <- c("code-3", "code-2", "lat", "lng", "name", "population")
countries$quant <- ntile(countries$population, 4)
colors <- data.frame(quant = c(1, 2, 3, 4),
                     color = c("#0033CC", "#003399", "#003355", "#003322"),
                     legend_text = c("25%", "50%", "75%", "100%"))
countries <- merge(countries, colors, by = "quant")

kable(head(countries, 4), digits=2)
```

Secondly, generate the map, and draw circles in the geographic country centers representing the absolute population (size), and the population quartile (color). By clicking the circle the a popup is shown with country name and population.
``` {r echo = FALSE, message = FALSE}
countries %>%
  leaflet() %>%
  addTiles() %>%
  addCircles(weight = 1,
             radius = sqrt(countries$population) * 50,
             popup = paste(countries$name, "-", prettyNum(countries$population,
                                                          big.mark = ",",
                                                          preserve.width = "none")),
             color = countries$color) %>%
  addLegend(labels = colors$legend_text, colors = colors$color)
```
